Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-21 _tex/paper2.tex
Built site for gh-pages

Researchers would then proceed to identify the estimands.
\emph{Identification} determines whether an estimator can accurately
compute an estimand based on its assumptions, regardless of random
variability \citep[pp.~4]{Schuessler_et_al_2023}. An \emph{estimator} is
a method or function that transforms data into an estimate
\citep{Neal_2020}. \emph{Estimates} are numerical values that
approximate the estimand, and they are derived through
\emph{estimation}, which refers to the process of integrating data with
an estimator \citep{Everitt_et_al_2010}. The Identification-Estimation
flowchart \citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow}
provides a visual representation of the process of transitioning from
estimands to estimates.

@@

While numerous methods can approximate an estimand, researchers
prioritize estimators with desirable identification properties that
ensure the accuracy of estimates. For instance, the Z-test is an
estimator known for its effectiveness when comparing groups'
proportions, yielding accurate estimates when the underlying assumptions
of the statistic are met \citep{Kanji_2006}. If this is the case, the
Z-test is expressed as a signal-to-noise statistic
\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is defined
as the difference between the groups' sample proportions,
\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
\(E[Y_{i} \:|\:T_{i}=1]\) and \(E[Y_{i} \:|\:T_{i}=2]\), respectively.
The noise, represented by \(\hat{s}_{p}\), is defined as the unpooled
sample variability observed between the two groups.

However, researchers often seek to uncover the mechanisms underlying
specific data and establish causal relationships rather than simply
estimate associations. In the example, researchers can interpret the
associational estimate represented by the Z-statistic as causal. This
interpretation relies on the data meeting the assumptions of the Z-test
and being collected through a randomized experiment.

@@

where \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \:|\:do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \:|\:do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Here, an
\emph{intervention} involves assigning a constant value to the treatment
variable for each student's potential outcomes. Note that if a student
is assigned to intervention \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
observable. To address this challenge, the structural approach extends
the ICE to the \emph{average causal effect} (ACE,
Equation~\ref{eq-ACE}), representing the average difference between
observed potential outcomes and their counterfactual counterparts.

@@

Even though counterfactuals are unobserved and unobservable, researchers
can still identify the ACE from associational estimates by leveraging
the structural approach. The approach identifies the ACE by
statistically conditioning data on a \emph{sufficient adjustment set} of
variables \(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.
This \emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones. If such a set
exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
(\(T \:\bot\:Y \:|\:X\)) \citep{Pearl_2009}, and \(X\) satisfies the
\emph{backdoor criterion} \citep[pp 37]{Neal_2020}. Here,
\emph{conditioning} describes the process of restricting the focus to
the subset of the population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).

@@

For instance, If researchers cannot conduct the randomized experiments
described in Section~\ref{sec-appendix-A} and must instead rely on
observational data, they can still identify the ACE as long as a
variable \(X\), such as the socio-economic status of the school,
satisfies the backdoor criterion. Under these circumstances, researchers
first identify the \emph{conditional average causal effect} (CACE,
Equation~\ref{eq-CACE})

@@

From the CACE, they identify the ACE from associational quantities as in
Equation~\ref{eq-mACE1}. This identification process is commonly known
as the \emph{backdoor adjustment}. Here, \(E_{X}[\cdot]\) represents the
marginal expected value over \(X\) \citep{Morgan_et_al_2014}.

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
models} that facilitate valid causal inference by guiding researchers in
determining which statistical models can identify an estimand (ACE,
CACE, or other), assuming the depicted causal structure is correct
\citep{McElreath_2020}. Figure~\ref{fig-IEflow} shows the role of
theoretical models in the inference process.

SCMs and DAGs support causal inference through two key advantages.
First, regardless of complexity, they can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.
Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
elaborate on the first advantage, while Section~\ref{sec-appendix-B22}
and Section~\ref{sec-appendix-B23} do so for the second.

\paragraph{SCMs and DAGs}\label{sec-appendix-B21}

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. The right panels of the
figures display these DAGs. A DAG is a graph consisting of nodes
connected by edges, where the nodes represent random variables. The term
\emph{directed} means that the edges extend from one node to another,
with arrows indicating the direction of causal influence. The term
\emph{acyclic} implies that the causal influences do not form loops,
ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. Although the \emph{standard representation}
of DAGs typically omits exogenous variables for simplicity, the
\emph{magnified representation} depicted in the figures offers some
advantages: including exogenous variables can help to highlight
potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

@@

A careful examination of the figures highlights the theoretical
assumptions underlying the observed variables in these building blocks.
SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} depict two unconnected
nodes, representing a scenario where variables \(X_{1}\) and \(X_{3}\)
are independent or not causally related. SCM \ref{fig-scm_bb2} and DAG
\ref{fig-mdag_bb2} illustrate two connected nodes, representing a
scenario where a \emph{parent} node \(X_{1}\) exerts a causal influence
on a \emph{child} node \(X_{3}\). In this setup, \(X_{3}\) is considered
a \emph{descendant} of \(X_{1}\). Additionally, \(X_{1}\) and \(X_{3}\)
are described as \emph{adjacent} because there is a \emph{direct path}
connecting them. SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict
a \emph{chain}, where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\)
influences \(X_{3}\). In this configuration, \(X_{1}\) is a parent node
of \(X_{2}\), which is a parent node of \(X_{3}\). This structure
creates a \emph{directed path} between \(X_{1}\) and \(X_{3}\).
Consequently, \(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and
\(X_{2}\) fully \emph{mediates} the relationship between the two. SCM
\ref{fig-scm_bb4} and DAG \ref{fig-mdag_bb4} illustrate a \emph{fork},
where variables \(X_{1}\) and \(X_{3}\) are both influenced by
\(X_{2}\). Here, \(X_{2}\) is a parent node that \emph{confounds} the
relationship between \(X_{1}\) and \(X_{3}\). Finally, SCM
\ref{fig-scm_bb5} and DAG \ref{fig-mdag_bb5} show a \emph{collider},
where variables \(X_{1}\) and \(X_{3}\) are concurrent causes of
\(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are not
causally related to each other but both influence \(X_{2}\) (an
\emph{inmorality}). Notably, all building blocks assume the errors are
independent of each other and from all other variables in the graph, as
evidenced by the pairwise relations \(e_{X1} \:\bot\:e_{X2}\),
\(e_{X1} \:\bot\:e_{X3}\), and \(e_{X2} \:\bot\:e_{X3}\).

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
DAG \ref{fig-mdag_example1} depict the plausible causal structure for
this example. In this context, the variable \(X\) (socio-economic status
of the school) acts as a confounder in the relationship between the
teaching method \(T\) and the outcome \(Y\). The figures display
multiple descendant relationships such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
unconnected node pairs, evident from the relationships
\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\paragraph{The probabilistic implications of SCMs and
DAGs}\label{sec-appendix-B22}

@@

\begin{equation}\phantomsection\label{eq-net_factor}{
P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \:|\:pa(X_{i}) )
}\end{equation}

@@

The five fundamental building blocks described in
Section~\ref{sec-appendix-B21} clearly illustrate which variables'
(in)dependencies can SMCs and DAGs predict. For instance, applying the
BNF to the causal structure shown in the SCM \ref{fig-scm_bb1} and DAG
\ref{fig-mdag_bb1} supports the theoretical assumption that the observed
variables \(X_{1}\) and \(X_{3}\) are unconditionally independent
(\(X_{1} \:\bot\:X_{3}\)). This conclusion follows from expressing their
joint probability distribution as
\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\) \citep[pp.~24]{Neal_2020}.
Conversely, when \(X_{3}\) is unconditionally dependent on \(X_{1}\)
(\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
distribution as \(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\).
Notably, these descriptions demonstrate the clear correspondence between
the structural equations illustrated in Section~\ref{sec-appendix-B21}
and the CPDs.

Beyond the insights gained from two-node structures, researchers can
uncover more nuanced patterns of variables' (in)dependencies from
chains, forks, and colliders. These (in)dependencies apply to any data
generated by these structures. For instance, applying the BNF to the
chain structure depicted in the SCM \ref{fig-scm_bb3} and DAG
\ref{fig-mdag_bb3} allow researchers to represent the joint distribution
for the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). This expression
implies that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
\((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\).
Moreover, in the fork structure shown in the SCM \ref{fig-scm_bb4} and
DAG \ref{fig-mdag_bb4}, researchers can express the joint distribution
of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1} \:|\:X_{2}) P(X_{2}) P(X_{3} \:|\:X_{2})\). This expression
allows researchers to further infer that \(X_{1}\) and \(X_{3}\) are
unconditionally dependent \((X_{1} \:\not\bot\:X_{3})\), but
conditionally independent when controlling for \(X_{2}\)
\((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). Finally, researchers analyzing the
collider structure illustrated in the SCM \ref{fig-scm_bb5} and DAG
\ref{fig-mdag_bb5} can express the joint distribution of the observed
variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}, X_{3}) P(X_{3})\). This representation
allows researchers to infer that \(X_{1}\) and \(X_{3}\) are
unconditionally independent \((X_{1} \:\bot\:X_{3})\), but conditionally
dependent when controlling for \(X_{2}\)
\((X_{1} \:\not\bot\:X_{3} \:|\:X_{2})\). The authors \citet[pp.~37, 40,
41]{Pearl_et_al_2016} and \citet[pp.~25--26]{Neal_2020} provide the
mathematical proofs for these conclusions.

Then, researchers can revisit the scenario in
Section~\ref{sec-appendix-B} using these insights. In this context,
applying the BNF to the SCM \ref{fig-scm_example2} and DAG
\ref{fig-mdag_example2} structures, can help researchers to represent
the joint probability of the observed variables as \(P(Y, T, X) =\)
\(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From this expression, researchers
can infer that the outcome \(Y\) is not unconditionally independent of
the teaching method \(T\) \((Y \:\bot\:T)\). This dependency arises from
two key structures: a direct causal path from the teaching method \(T\)
to the outcome \(Y\), represented by the two-connected-nodes structure
\(T \rightarrow Y\) (black path), and a confounding non-causal path from
the teaching method \(T\) to the outcome \(Y\) through the
socio-economic status of the school \(X\), represented by the fork
structure \(T \leftarrow X \rightarrow Y\) (gray path).

These probabilistic insights also indicate that narrowing the focus to a
subset of the population defined by the socio-economic status of the
school \(X\), achieved by conditioning on the variable and depicted by a
square in DAG \ref{fig-mdag_example2}, effectively blocks all
confounding non-causal path introduced by the fork structure.
Nevertheless, since \(Y\) remains dependent on \(T\) even after
conditioning \((Y \:\not\bot\:T \:|\:X)\), this dependency can only be
attributed to the direct causal effect \(T \rightarrow Y\).

@@

The structural approach to causal inference translate the probabilistic
insights from Section~\ref{sec-appendix-B23} into actionable strategies
for identifying the ACE from associational quantities. The approach
achieve this by relying on the modularity assumption and by re-defining
interventions in the form \(P(Y_{i} \:|\:do(T_{i}=t))\) (see
Section~\ref{sec-appendix-B}). The \emph{modularity assumption} posits
that intervening on a node alters only the causal mechanism of that
node, leaving others unchanged \citep[pp.~34]{Neal_2020}.

The modularity assumption underpins the concepts of manipulated graphs
and Truncated Factorization, which are essential for representing
interventions within SCMs and DAGs. \emph{Manipulated graphs} simulate
physical interventions by removing specific edges from a DAG, while
preserving the remaining structure unchanged \citep[pp.~34]{Neal_2020}.
In parallel, \emph{Truncated Factorization} (TF) achieves a similar
simulation by removing specific functions from the conceptual model and
replacing them with constants, while keeping the rest of the structure
unchanged \citep{Pearl_2010}. The probabilistic implications of this
factorization are formalized in Equation~\ref{eq-trunc_factor}, where
\(S\) represents the subset of variables \(X_{p}\) directly influenced
by the intervention.

@@

With these new insights, researchers revisiting the scenario in
Section~\ref{sec-appendix-B22} can infer that the socio-economic status
of the school, \(X\), satisfies the backdoor criterion, assuming the
causal structure depicted by the SCM \ref{fig-scm_example2} and DAG
\ref{fig-mdag_example2} is correct. This means that \(X\) serves as a
sufficient adjustment set, as it effectively blocks all confounding
non-causal paths introduced by the fork structure. In this context,
researchers can apply the \emph{backdoor adjustment} to identify the ACE
of \(T\) on \(Y\). They achieve this by first identifying the CACE of
\(T\) on \(Y\) by conditioning on \(X\), and then marginalizing this
effect over \(X\) to obtain the ACE, as expressed in
Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}). Notably, for
the purpose of identification, the conditioned DAG
\ref{fig-mdag_example2} is equivalent to the manipulated DAG
\ref{fig-mdag_example3}, because \(X\) satisfies the backdoor criterion.

@@

\subsection{The estimation process}\label{sec-appendix-C}

>>>>

Researchers would then proceed to identify the estimands.
\emph{Identification} determines whether an estimator can accurately
compute an estimand based on its assumptions, regardless of random
variability \citep[pp.~4]{Schuessler_et_al_2023}. An \emph{estimator} is
a method or function that transforms data into an estimate
\citep{Neal_2020}. \emph{Estimates} are numerical values that
approximate the estimand, and they are derived through
\emph{estimation}, referring to the process of integrating data with an
estimator \citep{Everitt_et_al_2010}. The Identification-Estimation
flowchart \citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow}
provides a visual representation of the process of transitioning from
estimands to estimates.

@@

While numerous estimators can approximate an estimand, researchers
prioritize estimators with desirable identification properties that
ensure the accuracy of estimates. For instance, the Z-test is an
estimator known for its effectiveness when comparing groups'
proportions, yielding accurate estimates when the underlying assumptions
of the statistic are met \citep{Kanji_2006}. If this is the case, the
Z-test is expressed as a signal-to-noise statistic
\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is defined
as the difference between the groups' sample proportions,
\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
\(E[Y_{i} \:|\:T_{i}=1]\) and \(E[Y_{i} \:|\:T_{i}=2]\), respectively.
The noise, represented by \(\hat{s}_{p}\), is defined as the unpooled
sample variability observed between the two groups.

However, researchers often seek to uncover the mechanisms underlying
specific data and establish causal relationships rather than simply
estimate associations. In the example, researchers can interpret the
associational estimand in Equation~\ref{eq-group_diff} and the
associational estimate represented by the Z-statistic as causal. This
interpretation is valid as long as the data satisfy the assumptions of
the Z-test and are collected through a randomized experiment.

@@

where \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \:|\:do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \:|\:do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Here, an
\emph{intervention} involves assigning a constant value to the treatment
variable for each student's potential outcomes. Note that if a student
is assigned to intervention \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
observable. To address this challenge, the structural approach extends
the ICE to the \emph{average causal effect} (ACE,
Equation~\ref{eq-ACE}), representing the average difference between the
students' observed potential outcomes and their counterfactual
counterparts.

@@

Even though counterfactuals are unobservable, researchers can still
identify the ACE from associational estimates by leveraging the
structural approach. The approach identifies the ACE by statistically
conditioning data on a \emph{sufficient adjustment set} of variables
\(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
\emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones. When such a set
exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
(\(T \:\bot\:Y \:|\:X\)) \citep{Pearl_2009}, and \(X\) satisfies the
\emph{backdoor criterion} \citep[pp 37]{Neal_2020}. Here,
\emph{conditioning} describes the process of restricting the focus to
the subset of the population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-group_diff}).

@@

For instance, If researchers cannot conduct the randomized experiments
described in Section~\ref{sec-appendix-A} and must instead rely on
observational data, they can still identify the ACE as long as an
observed variable \(X\), such as the socio-economic status of the
school, satisfies the backdoor criterion. Under these circumstances,
researchers first identify the \emph{conditional average causal effect}
(CACE, Equation~\ref{eq-CACE})

@@

From the CACE, researchers can identify the ACE from associational
quantities as in Equation~\ref{eq-mACE1}. This identification process is
commonly known as the \emph{backdoor adjustment}. Here, \(E_{X}[\cdot]\)
represents the marginal expected value over \(X\)
\citep{Morgan_et_al_2014}.

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
models} that facilitate valid causal inference. They achieve this by
helping researchers to determine which statistical models can identify
an estimand (ACE, CACE, or other), assuming the depicted causal
structure is correct \citep{McElreath_2020}. Figure~\ref{fig-IEflow}
shows the role of theoretical models in the inference process.

SCMs and DAGs support causal inference through two key advantages.
First, regardless of complexity, they can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.

Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
elaborate on the first advantage, while Section~\ref{sec-appendix-B22}
and Section~\ref{sec-appendix-B23} do so for the second.

\paragraph{The five fundamental block for SCMs and
DAGs}\label{sec-appendix-B21}

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. The right panels of the
figures display these DAGs. A DAG is a graph consisting of nodes
connected by edges, where the nodes represent random variables. The term
\emph{directed} means that the edges extend from one node to another,
with arrows indicating the direction of causal influence. The term
\emph{acyclic} implies that the causal influences do not form loops,
ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. Although the \emph{standard representation}
of DAGs typically omits exogenous variables for simplicity, the
\emph{magnified representation} depicted in the figures offers one key
advantage: including exogenous variables can help researchers highlight
potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

@@

A careful examination of these building blocks highlights the
theoretical assumptions underlying their observed variables. SCM
\ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} depict two unconnected
nodes, representing a scenario where variables \(X_{1}\) and \(X_{3}\)
are independent or not causally related. SCM \ref{fig-scm_bb2} and DAG
\ref{fig-mdag_bb2} illustrate two connected nodes, representing a
scenario where a \emph{parent} node \(X_{1}\) exerts a causal influence
on a \emph{child} node \(X_{3}\). In this setup, \(X_{3}\) is considered
a \emph{descendant} of \(X_{1}\). Additionally, \(X_{1}\) and \(X_{3}\)
are described as \emph{adjacent} because there is a \emph{direct path}
connecting them. SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict
a \emph{chain}, where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\)
influences \(X_{3}\). In this configuration, \(X_{1}\) is a parent node
of \(X_{2}\), which is a parent node of \(X_{3}\). This structure
creates a \emph{directed path} between \(X_{1}\) and \(X_{3}\).
Consequently, \(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and
\(X_{2}\) fully \emph{mediates} the relationship between the two. SCM
\ref{fig-scm_bb4} and DAG \ref{fig-mdag_bb4} illustrate a \emph{fork},
where variables \(X_{1}\) and \(X_{3}\) are both influenced by
\(X_{2}\). Here, \(X_{2}\) is a parent node that \emph{confounds} the
relationship between \(X_{1}\) and \(X_{3}\). Finally, SCM
\ref{fig-scm_bb5} and DAG \ref{fig-mdag_bb5} show a \emph{collider},
where variables \(X_{1}\) and \(X_{3}\) are concurrent causes of
\(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are not
causally related to each other but both influence \(X_{2}\) (an
\emph{inmorality}). Notably, all building blocks assume the errors are
independent of each other and from all other variables in the graph, as
evidenced by the pairwise relations \(e_{X1} \:\bot\:e_{X2}\),
\(e_{X1} \:\bot\:e_{X3}\), and \(e_{X2} \:\bot\:e_{X3}\).

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
DAG \ref{fig-mdag_example1} depict the plausible causal structure for
this example. In this context, the variable \(X\) (socio-economic status
of the school) is thought to be a confounder in the relationship between
the teaching method \(T\) and the outcome \(Y\). In this scenario, the
figures display multiple descendant relationships such as
\(X \rightarrow T\), \(X \rightarrow Y\), and \(T \rightarrow Y\). They
also highlight unconnected node pairs, evident from the relationships
\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\paragraph{The probabilistic implications of these
blocks}\label{sec-appendix-B22}

@@

\begin{equation}\phantomsection\label{eq-net_factor}{
\begin{aligned}
P(X_{1}, X_{2}, \dots, X_{P}) & = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \:|\:X_{i-1}, \dots, X_{1}) & (\small{\text{by chain rule})}\\
& = X_{1} \cdot \prod^{P}_{p=2} P(X_{i} \:|\:pa(X_{i}) ) & (\small{\text{by BNF}})
\end{aligned}
}\end{equation}

@@

The five fundamental building blocks described in
Section~\ref{sec-appendix-B21} clearly illustrate which (in)dependencies
can SMCs and DAGs predict. For instance, applying the BNF to the causal
structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
enables researchers to express the joint probability distribution of the
observed variables as \(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\),
supporting the theoretical assumption that the observed variables
\(X_{1}\) and \(X_{3}\) are unconditionally independent
(\(X_{1} \:\bot\:X_{3}\)) \citep[pp.~24]{Neal_2020}. Conversely, when
\(X_{3}\) is unconditionally dependent on \(X_{1}\)
(\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
distribution as \(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\).
Notably, these descriptions demonstrate the clear correspondence between
the structural equations illustrated in Section~\ref{sec-appendix-B21}
and the CPDs.

Beyond the insights gained from two-node structures, researchers can
uncover more nuanced patterns of(in)dependencies from chains, forks, and
colliders. These (in)dependencies apply to any data set generated by a
causal model with those structures, regardless of the specific functions
attached to the SCM \citep[pp.~36]{Pearl_et_al_2016}. For instance,
applying the BNF to the chain structure depicted in the SCM
\ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} allow researchers to
represent the joint distribution for the observed variables as
\(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}) P(X_{3} \:|\:X_{2})\). This expression
implies that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
\((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\).
Moreover, in the fork structure shown in the SCM \ref{fig-scm_bb4} and
DAG \ref{fig-mdag_bb4}, researchers can express the joint distribution
of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1} \:|\:X_{2}) P(X_{2}) P(X_{3} \:|\:X_{2})\). Similar to the
chain structure, this expression allows researchers to further infer
that \(X_{1}\) and \(X_{3}\) are unconditionally dependent
\((X_{1} \:\not\bot\:X_{3})\), but conditionally independent when
controlling for \(X_{2}\) \((X_{1} \:\bot\:X_{3} \:|\:X_{2})\). Finally,
researchers analyzing the collider structure illustrated in the SCM
\ref{fig-scm_bb5} and DAG \ref{fig-mdag_bb5} can express the joint
distribution of the observed variables as \(P(X_{1},X_{2},X_{3}) =\)
\(P(X_{1}) P(X_{2} \:|\:X_{1}, X_{3}) P(X_{3})\). This representation
allows researchers to infer that \(X_{1}\) and \(X_{3}\) are
unconditionally independent \((X_{1} \:\bot\:X_{3})\), but conditionally
dependent when controlling for \(X_{2}\)
\((X_{1} \:\not\bot\:X_{3} \:|\:X_{2})\). The authors \citet[pp.~37, 40,
41]{Pearl_et_al_2016} and \citet[pp.~25--26]{Neal_2020} provide the
mathematical proofs for these conclusions.

Using these additional probabilistic insights, researchers can revisit
the scenario in Section~\ref{sec-appendix-B}. In this context, applying
the BNF to the SCM \ref{fig-scm_example2} structure, enables the
representation of the joint probability distribution of the observed
variables as \(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From
this expression, researchers can infer that the outcome \(Y\) is
unconditionally dependent on the teaching method \(T\)
\((Y \:\not\bot\:T)\). This dependency arises from two key structures: a
direct causal path from the teaching method \(T\) to the outcome \(Y\),
represented by the two-connected-nodes structure \(T \rightarrow Y\)
(black path in DAG \ref{fig-mdag_example2}), and a confounding
non-causal path from the teaching method \(T\) to the outcome \(Y\)
through the socio-economic status of the school \(X\), represented by
the fork structure \(T \leftarrow X \rightarrow Y\) (gray path in DAG
\ref{fig-mdag_example2}).

@@

The structural approach to causal inference translate the probabilistic
insights from Section~\ref{sec-appendix-B23} into actionable strategies
seeking to identify the ACE from associational quantities. It achieves
this by relying on the \emph{modularity assumption}, which posits that
intervening on a node alters only the causal mechanism of that node,
leaving others unchanged \citep[pp.~34]{Neal_2020}.

The modularity assumption underpins the concepts of manipulated graphs
and Truncated Factorization, which are essential for representing
interventions \(P(Y_{i} \:|\:do(T_{i}=t))\) within SCMs and DAGs.
\emph{Manipulated graphs} simulate physical interventions by removing
specific edges from a DAG, while preserving the remaining structure
unchanged \citep[pp.~34]{Neal_2020}. In parallel, \emph{Truncated
Factorization} (TF) achieves a similar simulation by removing specific
functions from the conceptual model and replacing them with constants,
while keeping the rest of the structure unchanged \citep{Pearl_2010}.
The probabilistic implications of this factorization are formalized in
Equation~\ref{eq-trunc_factor}, where \(S\) represents the subset of
variables \(X_{p}\) directly influenced by the intervention, while an
example illustrating these concepts follows below.

@@

With these new insights, researchers revisiting the scenario in
Section~\ref{sec-appendix-B22} can infer that the socio-economic status
of the school, \(X\), satisfies the backdoor criterion, assuming the
causal structure depicted by the SCM \ref{fig-scm_example2} and DAG
\ref{fig-mdag_example2} is correct. This means that \(X\) serves as a
sufficient adjustment set, as it effectively blocks all confounding
non-causal paths introduced by the fork structure. Nevertheless, since
\(Y\) remains dependent on \(T\) even after conditioning
\((Y \:\not\bot\:T \:|\:X)\), this dependency can only be attributed to
the direct causal effect \(T \rightarrow Y\).

Researchers can then apply the \emph{backdoor adjustment} to identify
the ACE of \(T\) on \(Y\). They achieve this by first identifying the
CACE of \(T\) on \(Y\) by conditioning on \(X\), and then marginalizing
this effect over \(X\) to obtain the ACE. This process is expressed in
Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}). Notably, for
the purpose of identification, the conditioned DAG
\ref{fig-mdag_example2} is equivalent to the manipulated DAG
\ref{fig-mdag_example3}, because \(X\) satisfies the backdoor criterion.

@@

\paragraph{The estimation process}\label{sec-appendix-C}
